---
title: 'Reproducible Research - Peer Assignment #1'
author: "Bbuo"
output: html_document
---

This is an R markdown file to complete Peer Assignment #1. 

The first step is "Loading and preprocessing the data". We make sure the local directory is set explicitly. The file is input using read.csv function. 

```{r}
setwd("~/DATASCIENCE")
activityData <- read.csv('activity.csv')
str(activityData)
``` 

What is mean total number of steps taken per day?

Calculate the total number of steps taken per day. First, we need to split the data accordingly. 
Next, sum the appropriate subsets and produce a barplot. 

```{r}
stepsPerDay <- split(activityData$steps, activityData$date)
sumOfDay <- lapply(stepsPerDay, sum, na.rm = TRUE)
sumOfDay<- as.data.frame(sumOfDay)
table(sumOfDay)
```

Now, we produce a histogram of the data. 

```{r}
hist(as.numeric(sumOfDay))
```

Now, we want to get some basic stats like mean and median. 

```{r}
meanDays <- lapply(sumOfDay, mean, na.rm = TRUE)
medianDays <- lapply(sumOfDay, median, na.rm=TRUE)
table(meanDays)
table(medianDays)
```

What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
First, we do some data manipulation. 

```{r}
intervalData <- split(activityData$steps, activityData$interval)
intervalData <- as.data.frame(intervalData)
meanInterval <- as.numeric(apply(intervalData, 2,mean, na.rm = TRUE))
plot(meanInterval)
```

Now, we find the max location using: 

```{r}
maxInterval <- names(intervalData[which.max(meanInterval)])
```

which is: 

```{r}
maxInterval 
```

Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
Here, we use a simple sum: 

```{r}
numberOfNA <- sum(is.na(activityData$steps))
numberOfNA
```

Next, the strategy is to use the 5-min interval mean for the entire days.
Here, 

```{r}
activityData$interval2 <- c(1:288)

for (t in 1:length(activityData$steps)){
    if(is.na(activityData$steps[t])){
        activityData$steps[t] <- meanInterval[activityData$interval2[t]]
    }  
    else {
        activityData$steps[t] <- activityData$steps[t]
    }
}
```

And now copy the data into a new variable. 

```{r}
newActivityData <- activityData
activityData <- read.csv('activity.csv')
```

Now, we do some stats on the new data. 

```{r}
hist(newActivityData$steps)
```

Here, we can see that the data is pretty different to the first histogram. All of the NA are mostly averaged to near-zero values so the distribution has changed. 
Now we calculate some basic stats. 

```{r}
newDate <- split(newActivityData$steps, newActivityData$date)
meanNew <- lapply(newDate, mean, na.rm = TRUE)
medianNew <- lapply(newDate, median, na.rm=TRUE)
table(meanNew)
table(medianNew)
```

Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
newActivityData$date <- strptime(newActivityData$date, format="%Y-%m-%d", tz ="")
newActivityData$weekday <- weekdays(newActivityData$date)

ListWeekdays <- c("Monday", "Tuesday", "Wednesday","Thursday", "Friday")
ListWeekend <- c("Saturday","Sunday")

for (t in 1:length(newActivityData$date)){
    if(newActivityData$weekday[t] %in% ListWeekdays){
        newActivityData$weekday[t] <- "Weekday"
    }  
    else {
        newActivityData$weekday[t] <- "Weekend"
    }
}
```

Now, the new variable is produced: 

```{r}
head(newActivityData)
```

Next we analyze the average number of steps taken in weekdays and weekends. Therefore: 

```{r}
difData <- split(newActivityData, newActivityData$weekday)
mmmm <- difData[1]
llll <- difData[2]
weekdaysAVG <- split(mmmm$Weekday$steps, mmmm$Weekday$interval)
ccc <- as.numeric(lapply(weekdaysAVG, mean, na.rm = TRUE))
weekendAVG <- split(llll$Weekend$steps, llll$Weekend$interval)
ddd <- as.numeric(lapply(weekendAVG, mean, na.rm = TRUE))
```
And we plot the data using: 

```{r}
plot(ccc,ddd)
```